\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-30 tex/paper.tex
save

Complex computer simulators are increasingly used across fields of science to
describe generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely provide a way to directly evaluate the likelihood
function for a given observation. In this note, we develop a likelihood-free
inference algorithm for fitting a forward non-differentiable generative model to
(continuous or discrete) observed data. We adapt the adversarial training procedure of generative
adversarial networks by replacing the implicit generative network with a
domain-based scientific simulator, and solve the resulting non-differentiable
minimax problem by minimizing variational upper bounds of the adversarial objectives.
Effectively, the procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the corresponding
marginal distribution of the generated data matches the observations.
\glnote{Mention experimental results.}
\glnote{Add 'so what?' conclusion.}

@@

The observed data is sampled from a Poisson with scale $\lambda^* = 7$.
Algorithm~\ref{alg:avo} is run for 300 epochs with mini-batches of size $M=64$
and the following configuration. For the critic $d$, we use a 3-layer MLP with 10
hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
$n_\text{critic}=100$ iterations with a step size $\alpha=0.01$, decay rates
$\beta_1=\beta_2=0.5$ and its inner first and second moment vectors reset at
each outer epoch in order to avoid building momentum in staled directions.  For
estimating $\lambda^*$, we parameterize $\theta$ as $\log(\lambda)$ and use a univariate Gaussian proposal distribution
$q(\theta|\psi)$ initialized with a mean at $\log(5)$ and unit variance. At
each epoch, parameters $\psi$ are updated by taking one Adam step, with
$\alpha=0.01$ and $\beta_1=\beta_2=0.1$. The gradient penalty coefficient is set to
$\lambda=0.025$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=5$.

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
    \caption{Discrete Poisson model with unknown scale.
             ({\it Top left}) Proposal distributions $p(\theta|\psi)$ after adversarial variational optimization. For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
                        the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density and therefore to better inference.
             ({\it Top right}) Model distributions $p(\mathbf{x}|\psi)$ after training. This plot shows that the resulting parameterizations of the simulator closely reproduce the true distribution.
             ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
             }\label{fig:poisson}
\end{figure}

The top left plot in Figure~\ref{fig:poisson} illustrates the resulting proposal
distributions $p(\theta|\psi)$ after adversarial variational optimization.  For
both $\gamma=0$ and $\gamma=5$, the proposal distributions correctly concentrate
their density around the true parameter value $\log(\lambda^*) = 1.94$. Under
the effect of the positive entropy penalty $H(q(\theta|\psi))$,
the proposal distribution for $\gamma=5$ concentrates its mass more tightly,
yielding in this case a more precise inference.  The top right plot compares the
model distributions to the true distribution.  As theoretically expected from
adversarial training, we see that the resulting distributions closely match
the true distribution, with in this case slightly  better results for the penalized
model.  The bottom plot of Figure~\ref{fig:poisson} shows empirical estimates
of $-U_d$ with respect to the epoch number. For both $\gamma=0$ and $\gamma=5$,
the curves quickly fall towards $0$, which indicates that
$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)}
[d(\tilde{\mathbf{x}};\phi)] \approx \mathbb{E}_{\mathbf{x} \sim
p_r(\mathbf{x})} [d(\mathbf{x};\phi)]$ and that the critic cannot distinguish
between true and model data. Despite the discreteness and the
non-differentiability of the underlying generator, this confirms that inference
with adversarial variational optimization works.

@@

As a second example, we evaluate parameter inference for a generator producing
5-dimensional continuous data, as originally specified in Section 4.2. of
\citep{cranmer2015approximating}. More specifically, we consider the following
generative process:
\begin{itemize}
    \item $\mathbf{z} = (z_0, z_1, z_2, z_3, z_4)$, such that
        $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
        $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
        $z_2 \sim {\text{Mixture}}(\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5))$,
        $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
        $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
    \item $\mathbf{x} = R  \mathbf{z}$, where $R$ is a fixed
    semi-positive definite $5 \times 5$ matrix defining a fixed projection
    of $\mathbf{z}$ into the observed space.
\end{itemize}
We consider observed data generated at the nominal values $\theta^* = (\alpha^*=1,\beta^*=-1)$.
The simulator parameters are modeled with a factored Gaussian
proposal distribution $q(\theta|\psi) = q(\alpha|\psi) q(\beta|\psi)$, where each component is
initialized with zero mean and unit variance. The entropy penalty is set
to $\gamma=5$, while all other hyper-parameters are otherwise the same as in the previous example.

Starting with a proposal distribution $q(\theta|\psi)$ largely spread over
the parameter space, as illustrated in the left plot of Figure~\ref{fig:multi},
inference quickly converges towards a proposal distribution whose density
concentrates around the nominal values $\theta^*$, as shown in the right plot of Figure~\ref{fig:multi}.

@@

\subsection{Weinberg}

@@

\glnote{avo = extension of gans to non-differentiable generator}
\glnote{nice for discrete data}
\glnote{next version = try on an actual particle physics simulator.}

>>>>

Complex computer simulators are used across fields of science to
describe generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely provide a way to directly evaluate the likelihood
function for a given observation. In this note, we develop a likelihood-free
inference algorithm for fitting a forward non-differentiable generative model to
continuous or discrete observed data. We adapt the adversarial training procedure of generative
adversarial networks by replacing the implicit generative network with a
domain-based scientific simulator, and solve the resulting non-differentiable
minimax problem by minimizing variational upper bounds of the adversarial objectives.
Effectively, the procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the corresponding
marginal distribution of the generated data matches the observations.
Preliminary results with non-differentiable simulators producing discrete/continuous univariate/multi-dimensional data
validate the proposed method.

@@

The observed data is sampled from a Poisson with scale $\lambda^* = 7$.
Algorithm~\ref{alg:avo} is run for 300 epochs with mini-batches of size $M=64$
and the following configuration. For the critic $d$, we use a 3-layer MLP with 10
hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
$n_\text{critic}=100$ iterations with a step size $\alpha=0.01$, decay rates
$\beta_1=\beta_2=0.5$ and its inner first and second moment vectors reset at
each outer epoch in order to avoid building momentum in staled directions.  For
estimating $\lambda^*$, we parameterize $\theta$ as $\log(\lambda)$ and use a univariate Gaussian proposal distribution
$q(\theta|\psi)$ initialized with a mean at $\log(5)$ and unit variance. At
each epoch, parameters $\psi$ are updated by taking one Adam step, with
$\alpha=0.01$ and $\beta_1=\beta_2=0.5$. The gradient penalty coefficient is set to
$\lambda=0.025$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=5$.

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
    \caption{Discrete Poisson model with unknown scale.
             ({\it Top left}) Proposal distributions $q(\theta|\psi)$ after adversarial variational optimization. For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
                        the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density and therefore to better inference.
             ({\it Top right}) Model distributions $p(\mathbf{x}|\psi)$ after training. This plot shows that the resulting parameterizations of the simulator closely reproduce the true distribution.
             ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
             }\label{fig:poisson}
\end{figure}

The top left plot in Figure~\ref{fig:poisson} illustrates the resulting proposal
distributions $q(\theta|\psi)$ after adversarial variational optimization.  For
both $\gamma=0$ and $\gamma=5$, the proposal distributions correctly concentrate
their density around the true parameter value $\log(\lambda^*) = 1.94$. Under
the effect of the positive entropy penalty $H(q(\theta|\psi))$,
the proposal distribution for $\gamma=5$ concentrates its mass more tightly,
yielding in this case a more precise inference.  The top right plot compares the
model distributions to the true distribution.  As theoretically expected from
adversarial training, we see that the resulting distributions closely match
the true distribution, with in this case slightly  better results for the penalized
model.  The bottom plot of Figure~\ref{fig:poisson} shows empirical estimates
of $-U_d$ with respect to the epoch number. For both $\gamma=0$ and $\gamma=5$,
the curves quickly fall towards $0$, which indicates that
$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)}
[d(\tilde{\mathbf{x}};\phi)] \approx \mathbb{E}_{\mathbf{x} \sim
p_r(\mathbf{x})} [d(\mathbf{x};\phi)]$ and that the critic cannot distinguish
between true and model data. Despite the discreteness and the
non-differentiability of the underlying generator, this confirms that inference
with adversarial variational optimization works.

@@

As a second example, we evaluate parameter inference for a generator producing
5-dimensional continuous data, as originally specified in Section 4.2. of
\citep{cranmer2015approximating}. More specifically, we consider the following
generative process:
\begin{itemize}
    \item $\mathbf{z} = (z_0, z_1, z_2, z_3, z_4)$, such that
        $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
        $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
        $z_2 \sim {\text{Mixture}}(\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5))$,
        $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
        $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
    \item $\mathbf{x} = R  \mathbf{z}$, where $R$ is a fixed
    semi-positive definite $5 \times 5$ matrix defining a fixed projection
    of $\mathbf{z}$ into the observed space.
\end{itemize}
We consider observed data generated at the nominal values $\theta^* = (\alpha^*=1,\beta^*=-1)$.
The simulator parameters are modeled with a factored Gaussian
proposal distribution $q(\theta|\psi) = q(\alpha|\psi) q(\beta|\psi)$, where each component is
initialized with zero mean and unit variance. Hyper-parameters are set to $M=64$, $n_\text{critic}=100$, $\lambda=0.025$, $\gamma=10$ and
Adam configured with $\alpha=0.01$, $\beta_1=0.9$ and $\beta_2=0.999$.

Starting with a proposal distribution $q(\theta|\psi)$ largely spread over
the parameter space, as illustrated in the left plot of Figure~\ref{fig:multi},
inference quickly converges towards a proposal distribution whose density
concentrates around the nominal values $\theta^*$, as shown in the right plot of Figure~\ref{fig:multi}.
Overall, this example further illustrates and confirms the ability of adversarial
variational optimization for inference with multiple parameters and multidimensional
data, where reliable approximations of $p(\mathbf{x}|\theta)$ in a traditional
MLE setting would otherwise be difficult to construct.

@@

\subsection{Electron--positron annihilation}

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/weinberg.pdf}
    \caption{Electron--positron annihilation.
    ({\it Top left}) Proposal distributions $q(\theta|\psi)$ after adversarial variational optimization. The density of the penalized distribution ($\gamma=5$) is here highly concentrated, resulting in the green mass near $\theta^*$.
    ({\it Top right}) Model distributions $p(\mathbf{x}|\psi)$ after training. Despite the differences between their respective proposal distributions, both models closely match the observed data.
    ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
             }\label{fig:weinberg}
\end{figure}

As a more challenging example, we now consider a (simplified) simulator from
particle physics for electron--positron collisions resulting in muon--antimuon
pairs ($e^+e^- \rightarrow \mu^+\mu^-$). The simulator approximates the
distribution of observed measurements $\mathbf{x} = \cos(A) \in [-1,1]$, where $A$ is the
polar angle of the outgoing muon with respect  to the originally incoming
electron. This random variable is approximately distributed as
\begin{equation}
    p(\mathbf{x}|E_\text{beam}, G_f) = \frac{1}{Z} \left[ (1 + \mathbf{x}^2) + c(E_\text{beam}, G_f) \mathbf{x} \right]
\end{equation}
where $Z$ is a known normalization constant and $c$ is an asymmetry coefficient
function. Due to the linear term in the expression, the density $p(\mathbf{x} |
E_\text{beam}, G_f)$ exhibits a so-called {\it forward-backward} asymmetry.  Its
size depends on the values of the parameters $E_\text{beam}$ (the beam energy)
and $G_f$ (the Fermi constant) through the coefficient function $c$. Given the
known analytical form of the probability density function, a simulator for
$\mathbf{x}$ can be implemented with rejection sampling~\citep{von195113}. As an
arbitrary number of random drawings may be required before passing the
acceptance threshold, this illustrates how the latent variable
$\mathbf{z}$ does not necessarily need to be fixed-sized vector, and how its
density  $p(\mathbf{z}|\theta)$ may depend on $\theta$ in some complicated way
(Section~\ref{sec:problem}).

In this example, we consider observed data generated with $\theta^* =
(E_\text{beam}^*=42, G_f^*=0.9)$. The simulator parameters are modeled with a
factored Gaussian proposal distribution $q(\theta|\psi) = q(E_\text{beam}|\psi)
q(G_f|\psi)$, where each component is respectively initialized with mean $45$
and $1$ and variance $1$ and $0.01$. Hyper-parameters are set to $M=64$,
$n_\text{critic}=100$, $\lambda=0.0025$ and Adam configured with $\alpha=0.01$,
$\beta_1=0.9$ and $\beta_2=0.999$. As for the first example, we compare
entropy penalties $\gamma=0$ and $\gamma=5$.

The top left plot in Figure~\ref{fig:poisson} illustates the resulting proposal
distributions $q(\theta|\psi)$ for $\gamma=0$ and $\gamma=5$ after adversarial
variational optimization. We see that the distributions both arrive in the
neighborhood of $\theta^*$, with a density highly more concentrated for
$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
distance with $\theta^*$, both models closely match the observed data, as shown
in the top right plot of  Figure~\ref{fig:poisson}. This suggests either a
relatively flat landscape around $\theta^*$ or that the observed data can in
this case also be reproduced with a smeared variation $p(\mathbf{x}|\psi)$ of
the simulator. Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that
for both $\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly
fall towards $0$, which indicates  convergence towards a distribution that the
critic cannot distinguish from the true distribution.

@@

\glnote{avo = extension of gans to non-differentiable generator}
\glnote{nice for discrete data}
\glnote{next version = try on an actual particle physics simulator.}
\glnote{nes for better invariance and stability?}
